Figures to insert in text

### 2018-02-23 Chapters/Chapter02.tex
Many changes in SDM text

@@

@@

The framework implements the basic operations in a Sparse Distributed Memory which may be used to create more complex operations. It is developed in C language --- which may be loaded in many platforms and programming languages --- with a wrapper in Python. The Python module makes it easy to create and execute operations in a Sparse Distributed Memory.

We split the SDM memory in two parts: the hard-location address space and the hard-location counters. Thus, the addresses (bitstrings) of the hard-locations are stored in one array, while their counters in another. This may be useful to create multiple SDMs using the same address space, which would save computational effort to scan a bitstring in all the SDMs --- since they share the same address space, the activated hard-locations will be same in all of them. As the slowest part of reading and writing operations is scanning the address space, the performance benefits are high.

Each part may be stored either in the RAM memory or in a file. The RAM memory is interesting for quick experiments, automated tests, and others scenarios in which the SDM may be lost. The file is interesting for a long-term SDM --- one may create an SDM file with 10,000 random writes, which will be copied over and over to run multiple experiments. The file may also be sent to another researcher or may be published with the paper to let others run their own checks and verify the results. In summary, the framework fits many different uses and necessities.

There are some other advantages splitting the address space and the counters. When the scanner is run in the CPU, the accessed addresses in the RAM are continuous which improves considerably performance. When the scanner is run in the GPU, only the address space must be uploaded to the GPU, reducing the required memory.

Let a SDM memory with $N$ dimensions and $H$ hard-locations. Then, in a 64-bit computer, the array of hard-location addresses will use $H \cdot 8 \cdot \lceil N/64 \rceil$ bytes of memory, and there will be $H \cdot N$ hard-location counters. For example, in a SDM memory with 1,000 dimensions and 1,000,000 hard-locations, using 16-bit integers for the counters, the array of addresses will use 122MB of memory and the counters will use 1.9 GB of memory.

Basic operations were grouped in four sets: (i) for bitstrings, (ii) for addresses, (iii) for counters, and (iv) for memories (SDMs). Operations include creating new bitstrings, flipping bits, generating bitstring with a specific distance from a given bitstring, scan the address space using different algorithms, write a bitstring to a counter, write in an SDM, read from an SDM, and iteratively read from an SDM until convergence.

The reading and writing operations are executed in two steps: first, the address space is swept looking for the activated addresses; then, the operation is performed in the counter. Reading operation assemblies the bitstring according to the counters of the activated addresses, while the writing operation changes the counters.

Counters may have overflow protection depending on compiling options. By default, there is no overflow check in the counters, since each counter is a signed 32-bit integer --- which can store any number integer -2,147,483,648 and 2,147,483,647 --- and they will not overflow with less than $2^31-1$ divided by the number of activated hard-locations. When $N=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard-locations is 1,000 and it would require at least one million writes to overflow a counter.

\section{Implementation details}

@@

The hamming distance between two bitstrings is calculated counting the number of ones in the XOR of the bitstrings, i.e., $d(x, y) = \text{number of ones in} x \oplus y$. In order to count the number of ones of the XOR, the framework uses a pre-calculated table of distances --- by default, it is used a 16-bit table which uses 65MB of RAM, but one may increase to a 32-bit table which uses 4GB of RAM. Finally, to calculate the distance between two bitstrings, it sums the distance of each 16-bit part of the bitstrings, i.e., $d(x[0:N-1], y[0:N-1]) = d(x[0:15], y[0:15]) + d(x[16:31], y[16:31]) + \dots$. As each distance is calculated in O(1), this algorithm runs in O($\lceil bits/16 \rceil$).

The address space is stored in an array of addresses, which, in turn, is stored in arrays of 64-bit integers (Fig. \ref{tab:hl-addresses-detail}). As the addresses and their 64-bit integers are contiguous in RAM or file, there is a performance gain because processors load them in chunks.

@@

\caption{Hard-location addresses are stored in a contiguous array. In a 64-bit computer, each hard-location address is stored in a sub-array of 64-bit integers, with length $8 \cdot \lceil N/64 \rceil$.\label{tab:hl-addresses-detail}}
\end{figure}

@@

In 2014, there was published a solution to fast search in hamming space which seems applicable to our problem \citet{norouzi2014fast}.

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those algorithms is roughly to split the search space in half each step, which would take $O(\log(n))$ to go through the whole space. But this approach does not work in our case because of the high number of dimensions (i.e., 1,000) and because the hardlocations' addresses are randomly sampled from the ${0, 1}^d$ space. So, each addresses' bit itself splits the hardlocations in half, but it does not split the search space in half since both halves still must be convered. For instance, let's say we have 1,000 dimensions with 1,000,000 hardlocations, and we are scanning within a circle with radius 451, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. The search space has not been split in half because both halves have been covered (and one of them should have been skipped).

Hence, as our best approach approach seems to be scanning through all hardlocations, we may distribute the scan into many tasks which will be executed independently. The tasks may be executed in different processes, threads, or even computers. They may also run in the CPU or in the GPU. In this case, we may take into account both the time required to distribute the tasks and the time to receive their results.

@@

The OpenCL must be initialized, which just copies the hardlocations' addresses to the GPU's memory. Then, many scans may be executed with no necessity to upload the addresses again.

@@

\chapter{Appendix}
\section{Generating Kanerva's table 7.3}

\includepdf[
  pages=-,
  pagecommand={\pagestyle{headings}},
  %addtotoc={1,section,1,Quilling Shapes,sec:shapes}
]{./Chapter02-SDM/Kanerva-Table-7.3/Kanerva-Table-7_3.pdf}

>>>>

% !TEX root = ../partial-sdm.tex

@@

SDM memory or an extension of it has been used in many applications, helping to solve specific problems. \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs.

\citet{Linhares2011} showed that SDM respects the limits of short-term memory discussed by \citet{Miller1955} and \citet{Cowan2001}. This is the only work, to the author's knowledge, that explores some of intrinsic properties of SDM.

@@

The framework implements the basic operations in a Sparse Distributed Memory which may be used to create more complex operations. It is developed in C language --- which may be loaded in many platforms and programming languages --- with a wrapper in Python. The Python module makes it easy to create and execute simulations in a Sparse Distributed Memory and works properly in Jupyter Notebook \citep{kluyver2016jupyter}. It works in both Python 2 and Python 3.

We split the SDM memory in two parts: the hard-location addresses and the hard-location counters. Thus, the addresses (bitstrings) of the hard-locations are stored in one array, while their counters in another. This makes possible to create multiple SDMs using the same address space, which would save computational effort to scan a bitstring in all the SDMs --- since they share the same address space, the activated hard-locations will be same in all of them. As the slowest part of reading and writing operations is scanning the address space, the performance benefits are high.

Each part may be stored either in the RAM memory or in a file. The RAM memory is interesting for quick experiments, automated tests, and others scenarios in which the SDM may be lost, while the file is interesting for a long-term SDM, like creating an SDM file with 10,000 random writes, which will be copied over and over to run multiple experiments. The file may also be sent to another researcher or may be published with the paper to let others run their own checks and verify the results. In summary, the framework fits many different uses and necessities.

Let a SDM memory with $N$ dimensions and $H$ hard-locations. Then, in a 64-bit computer, the array of hard-location addresses will use $H \cdot 8 \cdot \lceil N/64 \rceil$ bytes of memory, and there will be $H \cdot N$ hard-location counters. For example, in a SDM memory with 1,000 dimensions and 1,000,000 hard-locations, using 32-bit integers for the counters, the array of addresses will use 122MB of memory and the counters will use 3.8 GB of memory.

Basic operations were grouped in four sets: (i) for bitstrings, (ii) for addresses, (iii) for counters, and (iv) for memories (SDMs). Operations include creating new bitstrings, flipping bits, generating bitstring with a specific distance from a given bitstring, scanning the address space using different algorithms, writing a bitstring to a counter, writing in an SDM, reading from an SDM, and iteratively reading from an SDM until convergence.

\section{Bitstring}

Bitstrings are the main structure of SDM. The addresses are represented in bitstrings, as well as the data. A bitstring is stored as an array of integers. Each integer may be 16-bit, 32-bit, or 64-bit long, depending on the configuration. By default, each integer is 64-bit long.

For instance, a 1,000-bit bitstring will have $\lceil 1000/64 \rceil = 16$ integers. These integers will have a total of $16 \cdot 64 = 1,024$ bits. The remaining 24 bits are always zero, so they do not affect the result of any operation. Bitstrings store neither how many bits they have nor the array length. These pieces of information are only stored in the address space.

@@

The distance between two bitstrings is calculated by the hamming distance, which is the number of different bits between them.

It is calculated counting the number of ones in the exclusive or (xor) between the bitstrings, i.e., $d(x, y) = \text{number of ones in} x \oplus y$. There are several algorithms to calculate the number of ones \citep{warren2013hacker}, but the performance depends on the processor. So, we have implemented three different algorithms and one may be selected through compiling flags. The default algorithm is to use a builtin instruction from the compiler.

An alternative algorithm available is the lookup. It pre-calculates a table of distances --- by default, it is a 16-bit table which uses 65MB of RAM, but one may increase to a 32-bit table which uses 4GB of RAM ---, which is accessed a few times. Finally, to calculate the distance between two bitstrings, it sums the distance of each 16-bit part of the bitstrings, i.e., $d(x[0:N-1], y[0:N-1]) = d(x[0:15], y[0:15]) + d(x[16:31], y[16:31]) + \dots$. As each distance is calculated in O(1), this algorithm runs in O($\lceil bits/16 \rceil$).

\section{Address space}

An address space is a fixed collection of bitstrings, and each bitstring represents a hard-location address. They store the number of bitstrings, as well as the number of bits, number of integers per bitstring, and the number of remaining bits.

Bitstrings are stored in a contiguous array of 64-bit integers, as shown in Figure \ref{tab:hl-addresses-detail}. Hence, there is some performance improvement when they are accessed \citep{pai2004linux}.

@@

\caption{Address space's bitstrings are stored in a contiguous array. In a 64-bit computer, each bitstring is stored in a sub-array of 64-bit integers, with length $8 \cdot \lceil N/64 \rceil$.\label{tab:hl-addresses-detail}}
\end{figure}

The scan for activated hard-locations is performed in an address space. It returns the indexes of the bitstrings which were inside the circle. Then, each operation uses these index in a different way.

@@

In 2014, there was published a solution to fast search in hamming space which seems applicable to our problem \citet{norouzi2014fast}. It provides a fast search when $r/b < 0.11$ or $r/b < 0.06$, where $r$ is the radius and $b$ is the number of bits. But, in our case, for a 1,000 bits SDM, $r/b = 0.451$, which changes the runtime to $O(n^{0.993})$. This is really close to $O(n)$, but with a larger constant. Hence, $O(n)$ is still faster.

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those computation geometry algorithms is roughly to split the search space in half each step, which would take $O(\log(n))$ to go through the whole space. But this approach does not work because of the high number of dimensions (i.e., 1,000) and because the hardlocations' addresses are randomly sampled from the ${0, 1}^d$ space. So, each addresses' bit itself splits the hardlocations in half, but it does not split the search space in half since both halves still must be convered by the algorithm. For instance, let's say we have 1,000 dimensions with 1,000,000 hardlocations, and we are scanning within a circle with radius 451, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. The search space has not been split in half because both halves have been covered (and one of them should have been skipped).

Hence, as our best approach is to scan through all hardlocations, we may distribute the scan into many tasks which will be executed independently. The tasks may be executed in different processes, threads, or even computers. They may also run in the CPU or in the GPU. In this case, we may take into account both the time required to distribute the tasks and the time to receive their results.

@@

The OpenCL must be initialized, which just copies the address space's bitstrings to the GPU's memory. Then, many scans may be executed with no necessity to upload the bitstrings again.

\section{Counters}

Each hard-location has one integer of data per bit. For instance, each hard-location of a 1,000 bits SDM has 1,000 bits. Those integers are stores in a counter.

A counter is an array of integers which stores the data of all hard-locations. So, the counter's array has $N \cdot H$ integers.

When two counters are added in a third counter, there may occur an overflow. Counters may have overflow protection depending on compiling options. By default, there is no overflow check for performance reasons. It is not supposed to be a problem because each counter is a signed 32-bit integer which can store any number between -2,147,483,648 and 2,147,483,647 (and they will not overflow with less than $2^{31}-1$ divided by the number of activated hard-locations). For instance, when $N=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard-locations is 1,000 and it would require at least one million writes to be possible to overflow a counter.

The reading and writing operations are executed in two steps: first, the address space is swept looking for the activated addresses; then, the operation is performed in the counters. Reading operation assemblies the bitstring according to the counters of the activated addresses, while the writing operation changes the counters.

The iterated reading keeps reading until it gets exactly the same bitstring (or the number of maximum interations has been reached), i.e., it performs $\eta_{i+1} = \text{read}(\eta_i)$ and stops when $\eta_{k+1} = \eta_{k}$. If the initial bitstring is inside the critical distance of $\eta$, it will converge to $\eta$, but, if it is not, it will diverge and reach the maximum number of iterations.

It is available both Kanerva's read and Chada's read. The latter was implemented according to a generalization proposed by professor Paulo Murilo (personal communication). The generalization has a $z$ parameters, which is the expoent of the counter. In this case, the results are floating point instead of integer, which reduces considerably the performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

There is also another special read operation: the weighted reading. The weight is retrieved from a lookup table of integers based on the distance between the hard-location's address and the reading address. The weight is multiplied by the counter's value and summed. The rest of the read operation is exactly the same.

@@

\chapter{Appendix}
%\section{Generating Kanerva's table 7.3}

%\includepdf[
%  pages=-,
%  pagecommand={\pagestyle{headings}},
%  %addtotoc={1,section,1,Quilling Shapes,sec:shapes}
%]{./Chapter02-SDM/Kanerva-Table-7.3/Kanerva-Table-7_3.pdf}

### 2018-02-23 Chapters/sdm2.tex
Many changes in SDM text

\section{Applications}

SDM memory or an extension of it has been used in many applications, helping to solve specific problems. \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs.

\citet{Linhares2011} showed that SDM respects the limits of short-term memory discussed by \citet{Miller1955} and \citet{Cowan2001}. This is the only work, to the author's knowledge, that explores some of intrinsic properties of SDM.

@@

@@

Starting from the premise of SDM as a faithful model of human short-term memory, a better understanding of the critical distance may shed light on our understanding of the thresholds that bind our own memory.

>>>>

@@

My advisor, professor Alexandre Linhares, has proposed another read operation: the weighted reading. In it, the sum of the counter's value is weighted based on the distance between the hard-location's address and the reading address. The logic behind it is to change the importance of each hard-location inside the circle. We have tested how SDM behaves when the weight is calculated using the information given by each bitstring, i.e., $w(d) = -\log_2 \left( 2^{-N} \binom{N}{d} \right) = N - \log_2 \binom{N}{d}$, as in information theory \citep{cover2012elements}.

@@

For further information about critical distance behavior, see \citet{Brogliato2011} and \citet{brogliato2014sparse}.
